---
title: "New Constuction Home Analysis"
author: "Nolan Peterson"
date: "1/8/2022"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, comment = NA)
```


## Introduction 

The aims of this analysis are:

* Observe and analyze the trends and relevant statistics for the new construction home market.

* Build a model to predict new home sale prices based off of selected features. 

## Data 

The data for this analysis comes from the Census Bureau's 2020 Survey of Construction. This is a survey of new construction homes across the US which compiles details regarding the construction methods used, physical specifications of the home as well as some legal/financial details surrounding the home. The data can be found here: https://www.census.gov/construction/chars/microdata.html

### Key Insights 

* Raw data has 25,526 observations and 61 variables. 

* Cleaned data has 12,443 observations and 39 variables.

* Cleaned data has 30 categorical predictors, 4 continuous predictors, 1 sample weight variable, and 1 continuous response. 

* Code for data cleaning can be found here: https://github.com/NolanPeterson453/New-Construction-Analysis-/blob/main/01_scripts/data_cleaning.R

### Cleaning Steps

#### Variable Selction

Not all variables in raw data set were useful for this analysis. There were primarily three reasons why variables would be removed. 

* The variable was a flag variable. 

* The variable had low or no predictive power. 

* The variable was proxy for the target response variable. 


For example, the variable SQFS_F was removed because this variable because this variable simply indicates if the Census Bureau has at all modified the reported value. Examples of modification to the reported value include top-coding, rounding or imputation. The second reason for removal (lack of predictive power), was used for variables such as ID or variables describing financing (loan type, etc...). These variables do not add any useful information to this analysis. The third type of variables removed, are variable which describe essentially the same thing thing as the response. That is they describe the value of the home in some measure.

#### Reported Vs Final Observations 

In the raw data set there are several variables which have both reported and 'final' reported counterparts. Take for example the variables SQFS (area of home in square feet) and FSQFS (final area of home in square feet). In cases where the home's final square footage area was different that what was specified in the building permit or contract, the final value is reported in the variable FSQFS. Otherwise the value of FSQFS is 0. In order to simplify the data while still maintaining the accuracy of the reported values the two variables SQFS and FSQFS where merged into a single variable which has the value of FSQFS when it is not 0 and the value of SQFS when FSQFS is 0. 

There is a similar situation for the response SLPR (sale price) and FSLPR (final sale price). FSLPR reports a value when the final sale price is different than what is specified in the contract and is 0 otherwise. These two variables were merged in the exact same way as SQFS and FSQFS. 

#### Missing Values 

There is a substantial amount of unreported values within the data set. For the categorical predictors missing data is simply coded in as an "unreported" or "non-applicable" level in the variable. For continuous variables, unreported values are imputed with the mean value of the reported observations. 

#### Observation Selection

Only observations where the home was sold were kept in the final data set. This is because sale price is the being used as the primary proxy for home value. 

### Training and Test Splits 

After all the above mentioned pre-processing steps were complete the data was split into a training data set and test data set. The training set is comprised of 80\% of the data while the remaining 20\% was devoted to the test set for model evaluation. 

## Exploritory Data Analysis 

### Key Insights 

* Distribution of home sale price is right skewed with the vast majority of homes selling for between \$125,000 and \$750,000. 

* There is a strong association between region of home and sale price. 

* There is a moderate association between size of home and sale price with interesting patterns to be explored. 

* Code relating to this section can be found here: https://github.com/NolanPeterson453/New-Construction-Analysis-/blob/main/01_scripts/explor_analysis_viz.R

### Sale Price of Home

THe sale price of the home is of particular interest because it can be used as a proxy of home value. It will be modeled as the response variable later on. For now though, there is a lot of information that can be gained with some simple analysis. The distribution of sale price is approximately normal with a right skew. 

```{r echo=FALSE}
# Library statements 
library(tidyverse)
library(splitstackshape)
library(hablar)
library(scales)
library(ggthemes)
library(ggrepel)
library(knitr)

# Read in data 
soc_data <- read_csv(file = "https://raw.githubusercontent.com/NolanPeterson453/New-Construction-Analysis-/main/02_data/cleaned_soc.csv")

# Assign column types 
factor_vars <- c("ACS", "AGER", "ASSOC", "BASE", "CON", "DECK", "DET", 
                 "DIV", "FNBS", "FOYER", "FRAME", "GAR", "HEAT", "HEAT2", 
                 "LNDR", "MFGS", "PATI", "PRCH", "SEWER", "STOR", "WAL1", 
                 "WAL2", "WALS","WATER", "BEDR", "FPLS", "FULB", "HAFB", "FUEL",
                 "FUEL2")

contin_vars <- c("FSLPR", "FSQFS", "LOTV", "FFNSQ", "AREA")

soc_data <- soc_data %>% 
  convert(fct(factor_vars),
          num(contin_vars))

# Expand by Sample weight 
soc_data_expanded <- expandRows(soc_data, 'WEIGHT')

# Summary statistics for response
summary(soc_data_expanded$FSLPR) 
cat(paste0('Mean Sale Price $',round(mean(soc_data_expanded$FSLPR), 2), '.\n'))
cat(paste0('Median Sale Price $',round(median(soc_data_expanded$FSLPR), 2), '.\n'))

# Distribution of Home Sale Price plot
cuts <- data.frame(ref = c("Mean Sale Price", "Median Sale Price"),
                   vals = c(mean(soc_data_expanded$FSLPR), median(soc_data_expanded$FSLPR)),
                   stringsAsFactors = FALSE)

soc_data_expanded %>% 
  ggplot(aes(x = FSLPR)) +
  geom_bar(stat="bin", 
           fill = "#244747") +
  scale_x_continuous(labels = dollar_format()) +
  scale_y_continuous(labels = comma) +
  labs(title = "Distribution of Home Sale Price",
       x = "Final Sale Price of Home", 
       y = "Count of Homes") + 
  geom_vline(aes(xintercept = vals, color = ref),
             data = cuts,
             show.legend = FALSE) +
  geom_text(aes(x = vals,
                y = c(50000,55000), 
                label = ref,
                color = ref,
                angle = 90, 
                vjust = 0.90),
            data = cuts,
            show.legend = FALSE)

```

#### Region of Home and Sale Price

It would not be illogical to assume that sale price depends the region which the home is located in. For example property values in west coast region are known to generally higher than the mid-west. This relationship can be visualized by using grouped box plots. As seen below the median sale price of homes in the East North Central, West North Central, South Atlantic, East South Central, and West South Central regions are roughly equal. While the median sale prices in the New England, Middle Atlantic, Mountain, and Pacific regions are all greater. 

```{r}
# custom color palette
econ_palette <- c("#e3120b", "#A37C27", "#91b8bd",
                  "#F05837", "#9ae5de", "#244747",
                  "#6A8A82", "#A7414A", "#660663")
# Boxplot of sale price grouped by region
regions <- c("New England",
             "Middle Atlantic",
             "East North Central",
             "West North Central",
             "South Atlantic",
             "East South Central",
             "West South Central",
             "Mountain",
             "Pacific")

soc_data_expanded %>% 
  ggplot(aes(y = FSLPR, x = DIV, color = DIV)) +
  geom_boxplot() +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_discrete(breaks = c(1:9),
                   labels = regions) +
  theme(axis.text.x = element_text(face="bold", color="black", 
                                   size=8, angle=90)) +
  labs(title = "Sale Price Grouped By Region",
       x = "Region",
       y = "Sale Price",
       color = "Region") +
  scale_color_manual(values = econ_palette, labels = regions)
```

To formally test weather an association exists between the sale price of the home and the region in which the home is located an analysis of variance (anova) can be conducted. The anova uses an F-test to test the following hypothesis:

H_0: There is no association between sale price and region. 

H_a: There is an association between sale price and region. 

```{r}
# Anova for association between DIV and FSLPR
ANOVA_FSLPR_DIV <- anova(lm(FSLPR ~ DIV, 
         data = soc_data,
         weights = WEIGHT))
ANOVA_FSLPR_DIV
```

As can be seen in the above anova table the p-value resulting from the F-test would result in the null hypothesis being rejected at the alpha = 0.001 level. **In other words there is a statistically significant association between sale price and the region of the home.** 

#### Size of Home and Sale Price 

The most obvious metric of home size is square footage. Using the correlation as a measure of association the r-squared value of 0.58 indicates a moderate association. 

```{r}
r_SQFS_FSLPR <- round(cor(soc_data_expanded$FSLPR, soc_data_expanded$FSQFS), 2)
cat(paste0('The correlation between sale price and square footage is ', r_SQFS_FSLPR, '.'))
```
Furthermore a linear model which predicts sale price using only square feet of home as a predictor in order to better understand the relationship between these two numbers. 

```{r}
# Linear Model 
FSLPR_FSQFS_mod <- lm(FSLPR ~ FSQFS, 
                     data = soc_data,
                     weights = WEIGHT)
summary(FSLPR_FSQFS_mod)
```


Using a T-test it is determined that square footage of a home is a significant predictor of sale price. It can also be shown from the linear model that for every one unit increase in square footage the sale price of the home is expected to rise by \$142.03. 

Other metrics of home size can also be useful predictors. For instance the number of bedrooms, bathrooms and half bathrooms are all variables which in addition to indicating the overall size of the house will provide information that can yield a more accurate estimate of sale price. 

```{r}
## Box plots for the sale price grouped by, bedrooms, bathrooms, and half baths
 
# Bedroom boxplot
bedroom_labs <- c("not reported",
                  "2 bedrooms or less",
                  "3 bedrooms",
                  "4 bedrooms",
                  "5 bedrooms or more")
bedr_plot <- soc_data_expanded %>% 
  ggplot(aes(y = FSLPR, x = BEDR, color = BEDR)) +
  geom_boxplot() +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_discrete(breaks = c(0, 2:5),
                   labels = bedroom_labs) +
  theme(axis.text.x = element_text(face="bold", color="black", 
                                   size=8, angle=90)) +
  labs(title = "Sale Price Grouped By Number of Bedrooms",
       x = "Bedrooms",
       y = "Sale Price",
       color = "Bedrooms") +
  scale_color_manual(values = econ_palette[1:5], labels = bedroom_labs)
bedr_plot

# Full bathroom boxplot 
fullbath_labs <- c("1 bathroom or less",
                   "2 bathrooms",
                   "3 bathrooms",
                   "4 bathrooms",
                   "not reported")
fulbath_plot <- soc_data_expanded %>% 
  ggplot(aes(y = FSLPR, x = FULB, color = FULB)) +
  geom_boxplot() +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_discrete(breaks = c(1:4, 9),
                   labels = fullbath_labs) +
  theme(axis.text.x = element_text(face="bold", color="black", 
                                   size=8, angle=90)) +
  labs(title = "Sale Price Grouped By Number of Full Bathrooms",
       x = "Full Bathrooms",
       y = "Sale Price",
       color = "Full Bathrooms") +
  scale_color_manual(values = econ_palette[1:5], labels = fullbath_labs)
fulbath_plot 

# Half bathroom boxplot
halfbath_labs <- c("0 half bathrooms",
                   "1 half bathroom",
                   "2 or more half bathrooms",
                   "not reported")
halfbath_plot <- soc_data_expanded %>% 
  ggplot(aes(y = FSLPR, x = HAFB, color = HAFB)) +
  geom_boxplot() +
  scale_y_continuous(labels = dollar_format()) +
  scale_x_discrete(breaks = c(0:2, 9),
                   labels = halfbath_labs) +
  theme(axis.text.x = element_text(face="bold", color="black", 
                                   size=8, angle=90)) +
  labs(title = "Sale Price Grouped By Number of Half Bathrooms",
       x = "Half Bathrooms",
       y = "Sale Price",
       color = "Half Bathrooms") +
  scale_color_manual(values = econ_palette[1:5], labels = halfbath_labs)
halfbath_plot

```

Interestingly homes with exactly 3 bedrooms have lowest median sale price when grouping sale price by the number of bedrooms a home has. The likely reason for this is that number of bedrooms is interacting with the location of the home. The data set does not specify if the home is located in a rural or urban area, however it would not be illogical to assume that 3 bedroom homes are the most common in rural areas where property values are low. This same behavior does not exist with with the number of bathrooms and half bathrooms. There is a high marginal cost associated with adding additional bathrooms to a home (due to the expensive fixtures and additional labor required). This cost must be reflected in the sale price of the home regardless of the location of the home. 


## Models 

Every model in the this section will be trained on the training set and evaluated on the test set. The main metric that will be used to evaluate each model is mean squared error of the model (MSE). Additionally the coefficient of determination (R-squared value) and the Akaike information criterion (AIC) will be reported or used in choosing the best linear regression model.  


### Multiple Linear Regression 

A full linear model was fit that used all predictors in the data set to estimate sale price of the home. Using step-wise selection with AIC as the criterion features were selected for either inclusion or removal in the model. The resulting model was used to predict on the test set. The MSE from the prediction on the test set was 2.0244e+10. The linear model performance acts as a baseline for comparison with future models. 

### Random Forest Models 





